Domain Specific Languages and Code Synthesis using Haskell

### 2014-03-14 paper.tex
Added the references

\documentclass[11pt]{article}
\usepackage{graphicx}
\usepackage{fancyvrb}
\usepackage{comment}
\usepackage[paper=letterpaper,margin=1.0in,includehead=false,includefoot=false]{geometry}

@@

\hyphenation{MATLAB} % prevent breaking over lines

@@

\title{Domain Specific Languages and Code Synthesis}
\author{Andy Gill, University of Kansas}
\maketitle

@@

There are many ways to give a computer instructions.
%
%% "might" might sound better than "may"
An electrical engineer might write a MATLAB program,
a database administrator may write an SQL script,
a hardware engineer may write in Verilog,
and an accountant may write a spreadsheet
with embedded formulas.
%
%% This sentence isn't clear, nor does it lead into the next sentence very well.
%% I'd suggest expanding on this point to make it clearer.
Aside from the difference in language, there is an
important difference in {\em form\/} and {\em idiom\/}.
%% Maybe you should emphasise "idiom" as well?
%
All of these examples use languages
customized to the job at hand, and build computational
requests in a form both familiar and productive
for the programmer (though an accountant may
not think of herself as a programmer.)
All these examples are uses of domain specific languages.

@@

There are two fundamental types of DSLs
(Figure~\ref{fig:types-of-dsls}).
%
The first, like the SQL example, is when a DSL is a first class language,
with its own compiler or interpreter, and is often used in
its own ecosystem. (Figure \ref{fig:types-of-dsls}(1)) 
All the examples mentioned so far fall
into this category. The primary difference between SQL and,
for instance, Java is one of scope and focus, though sometimes
DSLs grow to be as large as general purpose languages.
%
The other class of DSLs are languages
embedded in a host language(Figure \ref{fig:types-of-dsls}(2)). 
Such languages
can have the look and feel of being their own languages,
but leverage the host language's existing
ecosystem and initial semantics. 
(Figure \ref{fig:types-of-dsls}, (2))
It is this second
class of DSLs we are interested in, and is the subject of
this paper.

@@

An Embedded DSL (EDSL) is a language inside a language.
Haskell~\cite{Haskell98Book}, the premier pure functional programming language, is a great host for EDSLs
because of flexible overloading, a powerful type system, and lazy semantics facilitating this.
In this section, we give a terse
introduction to Haskell, 
sufficient to make this article self-contained. This section was
is an extended version the Haskell primer given by the author in~\cite{?}.

@@

There are many places this design pattern can be used. One example is the specification
of surface textures as functions; it is possible to export these into code executable
on GPUs, simultaneously lifting the abstractions used to write textures, and speeding
up how fast a implementation of the same operations would run.  There is nothing
that is specific about Haskell, or even functional languages here. Indeed, the same
ideas have been used in Java, for a VHDL generator\cite{...}. 
Haskell, with its powerful abstractions,
allows deep DSLs to almost feel like a straightforward shallow embedding.

@@

\end{Code}
Now, if we attempt to extract \verb|counter|, things go horribly wrong.
\begin{Code}
GHCi> counter (Var "restart") (Var "inc")
Mux2 (Var "inc") (Add (Mux2 (Var "restart") (Lit 0,Register 0 (Mux2 (Var "inc") ...
\end{Code}        
The output tree is infinite.
What has happened is the recursive definitions are unrolling when we try reify the function,
or more specifically, the body of counter is looping. At this point, the EDSL community was
stymied. There was efforts to use monadic structure, where the loop was expressing using
\verb|do|-notation~\cite{...}, making the loop an observable effect. There was an unsafe
extension to observe a limited form of sharing, by circumventing part of the purity of Haskell,
called observable sharing.
There was also an extension of Haskell input/output mechanism that allowed loops to
be observed indirect, called IO-based observable sharing. The net effect of all three
mechanisms is that the observed tree is rendered as a graph with named edges.

@@

\noindent
There are also extensions to the basic techniques. The principle once are:
\begin{itemize}
\item Internal function calls can captured as notes on a graph, 
rather than directly inlined~\cite{...}. 
This helps compilation of large programs, giving a basic separate compilation capability.
\item \verb|do|-statement can be reified, by normalization~\cite{...}. This result, called monadic
reification, is surprising. There are strong technical reasons to be believe monadic reification
should be impossible. However the normalization refactors the constrains that, by themselves would
be impossible to solve, and matches them up, 1-on-1, with a matching solution, allowing the whole
\verb|do|-notation to be solved and reified.
Monadic reification is a recent discovery, but has already been used in several Deep DSLs,
including Feldspar\cite{..} and Sunroof\cite{..}.
\item Control flow is problematic, and can not be used directly. However,
there is a generalization of Haskell boolean that do allow deep embedding capture~\cite{boolean}.
Using this library, a DSL with control flow can be constructed, but it needs to be explicit
code, at the DSL level, using constructors. The usage is clumsy, but workable, and we should
be able to do better.
\end{itemize}

@@

>>>>

\documentclass[11pt]{article}
\usepackage{graphicx}
\usepackage{fancyvrb}
\usepackage{comment}
\usepackage[paper=letterpaper,margin=1.0in,includehead=false,includefoot=false]{geometry}
\usepackage{cite}

@@

@@

\title{Domain Specific Languages and Code Synthesis using Haskell}
\author{Andy Gill, University of Kansas}
\maketitle

@@

There are many ways to give a computer instructions.
%
%% "might" might sound better than "may"
An electrical engineer might write a MATLAB program,
a database administrator may write an SQL script,
a hardware engineer may write in Verilog,
and an accountant may write a spreadsheet
with embedded formulas.
%
%% This sentence isn't clear, nor does it lead into the next sentence very well.
%% I'd suggest expanding on this point to make it clearer.
Aside from the difference in language, there is an
important difference in {\em form\/} and {\em idiom\/}.
%% Maybe you should emphasise "idiom" as well?
%
All of these examples use languages
customized to the job at hand, and build computational
requests in a form both familiar and productive
for the programmer (though an accountant may
not think of herself as a programmer.)
All these examples are use of domain specific languages.

@@

There are two fundamental types of DSLs
(Figure~\ref{fig:types-of-dsls}).
%
The first, like the SQL example, is when a DSL is a first class language,
with its own compiler or interpreter, and is often used in
its own ecosystem. (Figure \ref{fig:types-of-dsls}(1)) 
All the examples mentioned so far fall
into this category. The primary difference between SQL and
(say) Java is one of scope and focus, though sometimes
DSLs grow to be as large are general purpose languages.
%
The other class of DSLs are languages
embedded inside another hosting language(Figure \ref{fig:types-of-dsls}(2)). 
Such languages
can have the look and feel of being their own languages,
but leverage the host language to provide an existing
eco-structure and initial semantics. 
(Figure \ref{fig:types-of-dsls}, (2))
It is this second
class of DSLs we are interested in, and is the subject of
this paper.

@@

An Embedded DSL (EDSL) is a language inside a language.
Haskell~\cite{Haskell98Book}, the premier pure functional programming language, is a great host for EDSLs
because of flexible overloading, a powerful type system, and lazy semantics facilitating this.
In this section, we give a terse
introduction to Haskell, 
sufficient to make this article self-contained. This section was
is an extended version the Haskell primer given by the author in~\cite{Gill:11:Declarative}.

@@

There are many places this design pattern can be used. One example is the specification
of surface textures as functions; it is possible to export these into code executable
on GPUs, simultaneously lifting the abstractions used to write textures, and speeding
up how fast a implementation of the same operations would run.  There is nothing
that is specific about Haskell, or even functional languages here. Indeed, the same
ideas have been used in Java, for a VHDL generator\cite{Bellows:98:JHDL}. 
Haskell, with its powerful abstractions,
allows deep DSLs to almost feel like a straightforward shallow embedding.

@@

\end{Code}
Now, if we attempt to extract \verb|counter|, things go horribly wrong.
\begin{Code}
GHCi> counter (Var "restart") (Var "inc")
Mux2 (Var "inc") (Add (Mux2 (Var "restart") (Lit 0,Register 0 (Mux2 (Var "inc") ...
\end{Code}        
The output tree is infinite.
What has happened is the recursive definitions are unrolling when we try reify the function,
or more specifically, the body of counter is looping. At this point, the EDSL community was
stymied. There was efforts to use monadic structure, where the loop was expressing using
\verb|do|-notation~\cite{Erkok:2000:Recursive-Monad},
making the loop an observable effect. There was an unsafe
extension to observe a limited form of sharing, by circumventing part of the purity of Haskell,
called observable sharing~\cite{Claessen:99:ObserveSharing}.
There was also an extension of Haskell input/output mechanism that allowed loops to
be observed indirect, called IO-based observable sharing~\cite{Gill:09:TypeSafeReification}. 
The net effect of all three
mechanisms is that the observed tree is rendered as a graph with named edges.

@@

\noindent
There are also extensions to the basic techniques. The principle once are:
\begin{itemize}
\item Internal function calls can captured as notes on a graph, 
rather than directly inlined~\cite{Mainland:10:Nikola}. 
This helps compilation of large programs, giving a basic separate compilation capability.
\item \verb|do|-statement can be reified, by normalization~\cite{Sculthorpe:13:ConstrainedMonad,Svenningsson:13:CompositionalMonadReification,Persson:11:GenericMonadicConstructs}. This result, called monadic
reification, is surprising. There are strong technical reasons to be believe monadic reification
should be impossible. However the normalization refactors the constrains that, by themselves would
be impossible to solve, and matches them up, 1-on-1, with a matching solution, allowing the whole
\verb|do|-notation to be solved and reified.
Monadic reification is a recent discovery, but has already been used in several Deep DSLs,
including Feldspar\cite{Axelsson:2010:DandI-Feldspar} and Sunroof\cite{Bracker:14:Sunroof}.
\item Control flow is problematic, and can not be used directly. However,
there is a generalization of Haskell boolean that does allow deep embedding capture~\cite{hackage:boolean}.
Using this library, a DSL with control flow can be constructed, but it needs to be explicit
code, at the DSL level, using constructors. The usage is clumsy, but workable, and we should
be able to do better.
\end{itemize}

@@

\bibliography{paper}{}
\bibliographystyle{plain}
